---
title: "climate hack a thon"
author: "Amanda Bortner"
date: "2023-11-28"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readxl)
library(tidyverse)
library(ggplot2)
library(stringr)
```


```{r}
df <- read_excel("/Users/amandabortner/Desktop/Data Science Fundamentals/DS 3100 data/Prod_dataset.xlsx")
head(df)
```
State code represents the state the data is from 
the MSN is a code that specifies the type of renewable energy:
first two letters: energy source
second two letters: energy sector
last letter: type of data (P = physical unit, B = billion btu (british thermal unit))

first, going to remove all the rows that end in P because I have no way of consolidating those different units, and billion btu is a good standard unit to use as the measure, so going to remove all other ones

```{r}
df_btu <- df[str_ends(df$MSN, "B"), ]
```

Now going to group by states and sum the billions of btu along different energy sources and sectors to get a number for each state for each year 

```{r}
df_years <- df_btu %>%
  group_by(StateCode) %>%
  summarize(across(c(`1960`, `1962`, `1964`, `1966`, `1968`, `1970`, `1972`, `1974`, `1976`, `1978`, `1980`, `1982`, `1984`, `1986`, `1988`, `1990`, `1992`, `1994`, `1996`, `1998`, `2000`, `2002`, `2004`, `2006`, `2008`, `2010`, `2012`, `2014`, `2016`, `2018`, `2020`), sum, na.rm = TRUE))
```

removing rows that have a state code that aren't a US state code
```{r}
df_years <- df_years %>%
  filter(StateCode != "US") %>%
  filter(StateCode != "X3") %>%
  filter(StateCode != "X5")
```
there are now 51 rows, 50 states + DC, and an observation for each year and each state with the renewable energy production for energy sources and sectors that measure it in billion btu

```{r}
head(df_years) # checking it
```

Now going to split it up into Tennessee, Southeast, and USA

```{r}
df_TN <- df_years %>%
  filter(StateCode == "TN")
df_TN[, 1] <- NULL # removing state code
df_TN <- df_TN %>%
  pivot_longer(cols = everything(), 
               names_to = "year")


df_SE <- df_years %>%
  filter(StateCode %in% c("TN", "AL", "FL", "GA", "KY", "MS", "NC", "SC")) %>%
  summarize(across(c(`1960`, `1962`, `1964`, `1966`, `1968`, `1970`, `1972`, `1974`, `1976`, `1978`, `1980`, `1982`, `1984`, `1986`, `1988`, `1990`, `1992`, `1994`, `1996`, `1998`, `2000`, `2002`, `2004`, `2006`, `2008`, `2010`, `2012`, `2014`, `2016`, `2018`, `2020`), sum, na.rm = TRUE))
df_SE <- df_SE %>%
  pivot_longer(cols = everything(), 
               names_to = "year")


df_US <- df_years %>%
   summarize(across(c(`1960`, `1962`, `1964`, `1966`, `1968`, `1970`, `1972`, `1974`, `1976`, `1978`, `1980`, `1982`, `1984`, `1986`, `1988`, `1990`, `1992`, `1994`, `1996`, `1998`, `2000`, `2002`, `2004`, `2006`, `2008`, `2010`, `2012`, `2014`, `2016`, `2018`, `2020`), sum, na.rm = TRUE))
df_US <- df_US %>%
  pivot_longer(cols = everything(), 
               names_to = "year")
```


plotting lines 

```{r}
ggplot(df_TN, aes(x = year, y = value, group = 1)) +
  geom_point() +
  geom_line() +
  labs(x = "Time", y = "Value", title = "Line Graph of renewable energy production in TN (in BTU) over time") +
  theme_minimal()+
  scale_x_discrete(breaks = seq(min(df_TN$year), max(df_TN$year), by = 3))
```
```{r}
ggplot(df_SE, aes(x = year, y = value, group = 1)) +
  geom_point() +
  geom_line() +
  labs(x = "Time", y = "Value", title = "Line Graph of renewable energy production in the Southeast (in BTU) over time") +
  theme_minimal()+
  scale_x_discrete(breaks = seq(min(df_SE$year), max(df_SE$year), by = 3))
```
```{r}
ggplot(df_US, aes(x = year, y = value, group = 1)) +
  geom_point() +
  geom_line() +
  labs(x = "Time", y = "Value", title = "Line Graph of renewable energy production in the US (in BTU) over time") +
  theme_minimal()+
  scale_x_discrete(breaks = seq(min(df_US$year), max(df_US$year), by = 3))
```
```{r}
df_TN$Region <- "Tennessee"
df_SE$Region <- "Southeast US"
df_US$Region <- "United States"
```

```{r}
df_combined <- rbind(df_TN, df_SE, df_US)
```

```{r}
ggplot(df_combined, aes(x = year, y = value, color = Region)) +
  geom_point() +
  geom_line(aes(group = Region)) +
  labs(x = "Time", y = "Value", title = "Renewable Energy Production (in BTU) Comparison") +
  theme_minimal()+
  scale_x_discrete(breaks = seq(min(df_combined$year), max(df_combined$year), by = 3))
```
normalizing the values so they can be plotted together

```{r}
df_TN$normalized_values <- (df_TN$value - min(df_TN$value)) / (max(df_TN$value) - min(df_TN$value))

df_SE$normalized_values <- (df_SE$value - min(df_SE$value)) / (max(df_SE$value) - min(df_SE$value))

df_US$normalized_values <- (df_US$value - min(df_US$value)) / (max(df_US$value) - min(df_US$value))

df_combined <- rbind(df_TN, df_SE, df_US)

plot1 <- ggplot(df_combined, aes(x = year, y = normalized_values, color = Region)) +
  geom_point() +
  geom_line(aes(group = Region)) +
  labs(x = "Year", y = "Normalized Values of Renewable Energy Production", title = "Renewable Energy Production: Tennessee vs. the Southeast vs. the US") +
  theme_minimal()+
  scale_color_manual(values = c("Tennessee" = "#8DBEB3", "Southeast US" = "#6CA335", "United States" = "#9E681C"))+
  scale_x_discrete(breaks = seq(min(df_combined$year), max(df_combined$year), by = 3))
plot1
```


*** plot last 15 years anually

```{r}
df_last15 <- df_btu %>%
  group_by(StateCode) %>%
  summarize(across(c(`2006`, `2007`, `2008`, `2009`, `2010`, `2011`, `2012`, `2013`, `2014`, `2015`, `2016`, `2017`, `2018`, `2019`, `2020`, `2021`), sum, na.rm = TRUE))
```

removing rows that have a state code that aren't a US state code
```{r}
df_last15 <- df_last15 %>%
  filter(StateCode != "US") %>%
  filter(StateCode != "X3") %>%
  filter(StateCode != "X5")
head(df_last15)
```

```{r}
df_TNlast15 <- df_last15 %>%
  filter(StateCode == "TN")
df_TNlast15[, 1] <- NULL # removing state code
df_TNlast15 <- df_TNlast15 %>%
  pivot_longer(cols = everything(), 
               names_to = "year")

df_SElast15 <- df_last15 %>%
  filter(StateCode %in% c("TN", "AL", "FL", "GA", "KY", "MS", "NC", "SC")) %>%
 summarize(across(c(`2006`, `2007`, `2008`, `2009`, `2010`, `2011`, `2012`, `2013`, `2014`, `2015`, `2016`, `2017`, `2018`, `2019`, `2020`, `2021`), sum, na.rm = TRUE)) %>%
  pivot_longer(cols = everything(), 
               names_to = "year")

df_USlast15 <- df_last15 %>%
  summarize(across(c(`2006`, `2007`, `2008`, `2009`, `2010`, `2011`, `2012`, `2013`, `2014`, `2015`, `2016`, `2017`, `2018`, `2019`, `2020`, `2021`), sum, na.rm = TRUE)) %>%
  pivot_longer(cols = everything(), 
               names_to = "year")
```

```{r}
df_TNlast15$Region <- "Tennessee"
df_SElast15$Region <- "Southeast US"
df_USlast15$Region <- "United States"
```

```{r}
df_combinedlast15 <- rbind(df_TNlast15, df_SElast15, df_USlast15)
```

```{r}
ggplot(df_combinedlast15, aes(x = year, y = value, color = Region)) +
  geom_point() +
  geom_line(aes(group = Region)) +
  labs(x = "Time", y = "Value", title = "Renewable Energy Production (in BTU) Comparison") +
  theme_minimal()
```
normalizing the values so they can be plotted together

```{r}
df_TNlast15$normalized_values <- (df_TNlast15$value - min(df_TNlast15$value)) / (max(df_TNlast15$value) - min(df_TNlast15$value))

df_SElast15$normalized_values <- (df_SElast15$value - min(df_SElast15$value)) / (max(df_SElast15$value) - min(df_SElast15$value))

df_USlast15$normalized_values <- (df_USlast15$value - min(df_USlast15$value)) / (max(df_USlast15$value) - min(df_USlast15$value))

df_combinedlast15 <- rbind(df_TNlast15, df_SElast15, df_USlast15)
```

```{r}
install.packages("extrafont")
library(extrafont)

```

```{r}
#lighter gridlines
#annotation about what it shows and source of the data
plot2 <- ggplot(df_combinedlast15, aes(x = year, y = normalized_values, color = Region)) +
  geom_line(aes(group = Region)) +
  geom_point(shape = 21, aes(fill = Region), size = 2, color = "white") + 
  labs(x = "Year", y = "Normalized Values of Renewable Energy Production\nOriginally in BTU (British Thermal Units)", title = "Renewable Energy Production: Tennessee vs. the Southeast vs. the US") +
  theme_minimal() +
  theme(panel.grid.major = element_line(color = "grey96", size = 0.5),
        panel.background = element_rect(fill = "white", colour = "black"),
        plot.background = element_rect(fill = "white", colour = NA),
        text = element_text(family = "Avenir"),
        plot.title = element_text(size = 16)) +
  scale_color_manual(values = c("Tennessee" = "#8DBEB3", "Southeast US" = "#6CA335", "United States" = "#9E681C")) +
  scale_fill_manual(values = c("Tennessee" = "#8DBEB3", "Southeast US" = "#6CA335", "United States" = "#9E681C"))

plot2
```


```{r}
ggplot2::ggsave(
  filename = "EELU_data_viz.png", # change extension as needed
  plot = plot2, # change to your {ggplot2} object
  height = 6, # figure height, adjust as necessary
  width = 9, # figure width, adjust as necessary
  dpi = 1000 # high quality dots per inch 
)


```

1. draw a topic and suggestions for public data sources
2. create table/figure in software of choice
3. apply report's template style
4. submit by December 1, 2023

source: https://www.eia.gov/state/seds/seds-data-complete.php#Production (US EIA (Energy Information Administration)) data set of all production estimates

